The \dealii{} Library, Version 9.4

### 2022-04-24 9.4/paper.tex
Fix typos

\dealii{} version 9.4.0 was released June TODO, 2022.
This paper provides an
overview of the new features of this release and serves as a citable
reference for the \dealii{} software library version 9.4. \dealii{} is an
object-oriented finite element library used around the world in the
development of finite element solvers. It is available for free under the
GNU Lesser General Public License (LGPL). Downloads are available at
\url{https://www.dealii.org/} and \url{https://github.com/dealii/dealii}.

The major changes of this release are:
%
\begin{itemize}
  \item Advances in simplex- and mixed-mesh support (see Section~\ref{sec:simplex});
  \item Repartitioning of distributed meshes (see Section~\ref{sec:repartitioning});
  \item Advances in matrix-free infrastructure (see Section~\ref{sec:mf});
  \item Advances in multigrid infrastructure (see Section~\ref{sec:multigrid});
  \item CutFEM support (see Section~\ref{sec:cut});
  \item Performance improvement of particle infrastructure (see Section~\ref{sec:particles});
  \item Two new tutorial programs and a new code gallery program (see Section~\ref{subsec:steps}).
\end{itemize}
%

While all of these major changes are discussed in detail in
Section~\ref{sec:major}, there
are a number of other noteworthy changes in the current \dealii{} release
that we briefly outline in the remainder of this section:
%
\begin{itemize}
%  \item \texttt{AffineConstraints::make\_consistent\_in\_parallel()} allows
%  to make constraints consistent in parallel.
  \item \texttt{DataOutResample} does not output a numerical solution 
  on the cells of the original triangulation but interpolates the result 
  onto a second triangulation (that can be completely unrelated). 
  By using this class, one can output the result obtained on an 
  unstructured mesh on a structured one, which might be a more
  memory-efficient storage format, or one can create a slice in 3D.
  \item The new function \texttt{find\_point\_owner\_rank()} of \texttt{parallel::distributed::Triangulation} allows to find the MPI
  rank of the subdomain of a distributed mesh that contains a given point.
  It is communication-free, leverages the functionality of p4est (>v.2.2),
  and its algorithm is described in \cite{burstedde2020parallel}. Based on the obtained information,
  communication pattern in \texttt{Utilties::MPI::RemotePointEvaluation} can be set up efficiently. Furthermore, this feature could be used in the future to allow
  particle simulations, where particle movement is not
  limited by CFL conditions, as done in \cite{mirzadeh2016parallel}.
\end{itemize}
%
The changelog lists more than 150 other features and bugfixes.

@@

This release of \dealii{} contains a number of large and significant changes
that will be discussed in this section.
It of course also contains a
vast number of smaller changes and added functionality; the details of these
can be found
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_3_0_and_9_4_0.html}
{in the file that lists all changes for this release}; see \cite{changes94}.

@@

We continued to work on the simplex- and mixed-mesh support. We fixed
many bugs and generalized existing functions that only worked for
hypercube shaped cells. Most notable new functions are:

@@

  \caption{New refinement strategies for triangles and tetrahedrons: triangles
  are subdivided in 4 children and tetrahedrons 8 ones.}\label{fig:refinement}

@@

\begin{itemize}
\item Experimental support of locally refined meshes. FEM on locally refined
meshes requires 1) the possibility to locally refine the mesh (see Figure~\ref{fig:refinement})
and 2) appropriate hanging-node constraints. For 3D, we are in progress to
implement the missing constraint definitions.
\item TODO
\end{itemize}

Furthermore, we continued to remove the usage of  \texttt{GeometryInfo} from
the library and to replace the instances by more general equivalent functions, which
in many cases rely of \texttt{ReferenceCell}. Once all instance of 
\texttt{GeometryInfo} are remove, we will deprecate the class.

@@

Until now, distributed meshes were either partitioned statically (\texttt{parallel::\allowbreak fullydistributed::\allowbreak Triangulation}; short \texttt{p::f::T}) or used a
fixed policy to partition the cells among processes
(\texttt{parallel::\allowbreak distributed::\allowbreak Triangulation}; short \texttt{p::d::T}). For latter, we
use a Morton-order as the backend of that class---\texttt{p4est}---does. Fixed Morton-order
might provide very good performance in terms in communication and
setup costs, but might be non-optimal when interfacing with other
libraries that have fixed partitioning, most notably a Cartesian
partitioning, themselves.

New utility functions from the \texttt{RepartitioningPolicyTools} can be
used now to create a new \texttt{p::f::T} instances,
given a distributed triangulation (\texttt{p::f::T} or \texttt{p::d::T}) and a vector with the new owners of locally
owned cells. The workflow is shown in the following listing:
\begin{c++}
// original triangulation
parallel::distributed::Triangulation<dim> tria(comm);

@@

// create the new triangulation
parallel::fullydistributed::Triangulation<dim> tria_pft(comm);
tria_pft.create_triangulation(construction_data);
\end{c++}
Instead of using one of the predefined partitioning policies, users
can write their own by implementing the \texttt{RepartitioningPolicyTools::Base} interface. Similarly to the
active level, also the multigrid levels can be repartitioned arbitrarily.

\begin{figure}
    \centering
    \def\svgwidth{0.8\columnwidth}
    \input{svg/repartitioning.pdf_tex}
    \caption{Visualization of the repartitioning process: after 
    cells are assigned new ranks, each process collect sends the cells incl. 
    ghost cells to the new owner. There incoming cells are processed, duplicates
    are removed and the local part of the triangulation is built.}\label{fig:repartitioning}
\end{figure}

The setup process is visualized in Figure~\ref{fig:repartitioning}. First cells, incl. their
surrounding (ghost) cells and their parent cells, are collected and 
sent to the new owner. On the
receiving side, the set of all cells are combined and possible duplicates
are removed. This information is enough to construct a new triangulation.
For sending/receiving, we apply consensus-based algorithms~\cite{hoefler2010scalable}, which
we introduced in the library in release 9.2~\cite{dealII92}. Consensus-based 
algorithms are used also in \cite{ibanez2016pumi} for repartitioning.

In future releases, we plan to add support for repartitioning based on distributed graph
partitioning libraries, e.g., \texttt{ParMETIS} of \texttt{Zotan}. Furthermore, we intend to extend the 
\texttt{p::f::T} with adaptivity support.
With this and and the new repartitionging features, \texttt{p::f::T}
could be selfconsistent like \texttt{p::d::T}, which is an important
step towards the support of AMR also for distributed simplex and mixed meshes
(see also Subsection~\ref{sec:simplex}).

@@

In the matrix-free infrastructure, we have added many new features. Most
notably:
\begin{itemize}
\item Improved support of Hessians: just like values and gradients can be
evaluated and integrated during matrix-free loops, one can do this now
also for Hessians both for cells and faces. This enables, e.g., to write
a matrix-free version of step-47, which solves the biharmonic equation with DG.
\item Cell-centric loops now also allow to access to gradients and Hessians
of neighboring cells on faces. The major obstacle here was to reorientation
of theses in the case of arbitrary orientated cells on unstructured meshes.
\item Users can now create their own cell batches, by providing \texttt{FEEvaluation::reinit()} a list of cell IDs. \texttt{FEEvaluation}
accesses the appropriate data and reshuffles mapping data accordingly on
the fly to enable vectorization over cells. The new feature is useful in many
contexts. Examples are application with sharp interfaces (e.g., two phase flow
or shock capturing) where one needs to treat cells that are ``cut'' by
the interface in a special way. A challenge is that cell batchs
might contain cut or not-cut cell, making vectorization potentially more complication. One way to deal with such cell batches it to apply masks if
the code paths do not diverge too much
(many functions of \texttt{FEEvaluation} allows this). Another way is to
categorize cells during \texttt{MatrixFree::reinit()} in such a way that mixed cell batches do not occur. However, calling  \texttt{MatrixFree::reinit()}
might be too expensive if a very dynamic system is given, which requires
recatorization in each time step. Doing this on-the-fly might be a cheep alternative, even if not computationally optimal during matrix-free loops.
\end{itemize}
Besides these new features, we improved the performance of 
hanging-node-constraint evaluation on the CPU. Instead of performing
quasi-dense matrix-vector multiplications~\cite{KronbichlerKormann2012}, we now use an
approach based on inplace interpolation and sumfactorization, similarly
as we already did in the GPU code~\cite{ljungkvist2017matrix}. The algorithm is described
in \cite{munch2022hn} and performance numbers are shown, indicating a reduction
of overhead of cells with hanging nodes by a factor of 10$\times$.

Furthermore, we have performed a major restructuring of internals
of the \texttt{FEEvaluation} classes. This will enable us to add more
simply the support of new element types in the future, e.g., of 
Raviart--Thomas and Nedelec elements.

@@

In release 9.3~\cite{dealII93}, we added support for global coarsening multigrid in 
addition to the established local smoothing infrastructure. Global 
coarsening smoothens on the whole computational domain on each
multigrid level, which are obtained by coarsening the finest mesh globally.
For this purpose, we use a sequence of triangulations, on which we perform
the smoothing only on the active levels. To create the sequence of 
triangulations, one can use the functions \texttt{create\_geometric\_coarsening\_sequence()} from the
\texttt{MGTransferGlobalCoarseningTools} namespace. A new version takes an 
instance of \texttt{RepartitioningPolicyTools::Base} (see Subsection~\ref{sec:repartitioning}, which allow to specify the parallel 
distribution of each multigrid level (in contrast to the fixed---first-child---policy in the case of local smoothing). Furthermore, we added support of block vectors,
fixed a number of limitations, and performed performance optimizations of 
the transfer operator; particularly, the redundant copy from/to temporary vectors
has been eliminated and the application of hanging-node constraints are applied
efficiently the same way as in the matrix-free loops (see Subsection~\ref{sec:mf}).

In \cite{munch2022gc}, the performance of \texttt{deal.II}'s local-smoothing and global-coarsening
infrastructure was compared for locally refined meshes. The results indicate that
the local definition of multirgrid levels might introduce load imbalances
in the case of local smoothing so that global coarsening is favorable despite
of potentially more expensive inter-grid transfers. To judge the benefit
of one approach against the other, \texttt{deal.II} provides new functions
\texttt{workload\_imbalance()} and \texttt{vertical\_communication\_efficiency()} 
in the \texttt{MGTools} namespace to estimate the workload imbalance and the
vertical communication efficiency purely based on the given mesh. The publication also points out that not
all types of smoothers are applicable for global coarsening, due to the
presence of hanging-nodes, which is a motivation to add new smoother types
to \dealii.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{New and improved tutorials and code gallery programs}
\label{subsec:steps}

@@

The 9.4 release includes
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_3_0_and_9_4_0.html}
{around 40 incompatible changes}; see \cite{changes94}. The majority of these changes
should not be visible to typical user codes; some remove previously
deprecated classes and functions; and the majority change internal
interfaces that are not usually used in external
applications. That said, the following are worth mentioning since they
may have been more widely used:
\begin{itemize}
  \item In continuation of our attempt to merge the classes \texttt{DoFHandler} and \texttt{hp::DoFHandler}, we have removed the
  template parameter \texttt{DoFHandlerType} form many classes and
  functions and replaced it by \texttt{dim}/\texttt{spacedim}. Affected
  classes are, e.g., \texttt{SolutionTransfer} and \texttt{DataOut}. 
  \item \texttt{FE\_RaviartThomasNodal} now uses a different polynomial space to allow
for a simpler use for faces in non-standard orientation. The new polynomials
are anisotropic tensor products of Lagrange polynomials on the points of a
Gauss--Lobatto quadrature formula. This change leads to different entries in
the matrices and constraints, for example, but as the resulting polynomial
space spans the same polynomials, no change in accuracy should be expected.
\item We made different changes to the repartitoning of \texttt{parallel::distributed::Triangulation}. Most notably, we have have
removed the default weight (10,000) for consistency reasons with the rest
of the library.

>>>>

\dealii{} version 9.4.0 was released June 10, 2022.
This paper provides an
overview of the new features of this release and serves as a citable
reference for the \dealii{} software library version 9.4. \dealii{} is an
object-oriented finite element library used around the world in the
development of finite element solvers. It is available for free under the
GNU Lesser General Public License (LGPL). Downloads are available at
\url{https://www.dealii.org/} and \url{https://github.com/dealii/dealii}.

The major changes of this release are:
%
\begin{itemize}
  \item Advances in simplex- and mixed-mesh support (see Section~\ref{sec:simplex});
  \item Repartitioning of distributed meshes (see Section~\ref{sec:repartitioning});
  \item Advances in matrix-free infrastructure (see Section~\ref{sec:mf});
  \item Advances in multigrid infrastructure (see Section~\ref{sec:multigrid});
  \item CutFEM support (see Section~\ref{sec:cut});
  \item Performance improvement of particle infrastructure (see Section~\ref{sec:particles});
  \item Two new tutorial programs (see Section~\ref{subsec:steps}).
\end{itemize}
%

While all of these major changes are discussed in detail in
Section~\ref{sec:major}, there
are a number of other noteworthy changes in the current \dealii{} release,
which we briefly outline in the remainder of this section:
%
\begin{itemize}
%  \item \texttt{AffineConstraints::make\_consistent\_in\_parallel()} allows
%  to make constraints consistent in parallel.
  \item The \texttt{DataOutResample} class does not output a numerical solution 
  on the cells of the original triangulation, but interpolates the result 
  onto a second triangulation (which can be completely unrelated). 
  By using this class, one can output the result obtained on an 
  unstructured mesh on a structured one, which might be a more
  memory-efficient storage format, or one can create a slice in 3D.
  \item The new function \texttt{find\_point\_owner\_rank()} of \texttt{parallel::distributed::Triangulation} allows to find the MPI
  rank of the subdomain of a distributed mesh that contains a given point.
  It is communication-free and leverages the functionality of p4est (>v.2.2).
  Its algorithm is described in \cite{burstedde2020parallel}. Based on the information obtained,
  communication pattern in \texttt{Utilties::MPI::RemotePointEvaluation} can be set up efficiently. Furthermore, this function could be used in the future to allow
  particle simulations where particle movement is not
  limited by CFL conditions, as done in \cite{mirzadeh2016parallel}.
\end{itemize}
%
The changelog lists more than 150 other features and bugfixes.

@@

This release of \dealii{} contains a number of large and significant changes,
which will be discussed in this section.
It of course also includes a
vast number of smaller changes and added functionality; the details of these
can be found
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_3_0_and_9_4_0.html}
{in the file that lists all changes for this release}; see \cite{changes94}.

@@

We continued to work on the simplex- and mixed-mesh support. We fixed
many bugs and generalized existing functions that only worked for
hypercube-shaped cells. The most notable new functions are:

@@

  \caption{New refinement strategies for triangles and tetrahedrons: triangles
  are subdivided in 4 children and tetrahedrons in 8 ones.}\label{fig:refinement}

@@

\begin{itemize}
\item Experimental support of locally refined meshes. FEM on locally refined
meshes requires 1) the possibility to locally refine the mesh (see Figure~\ref{fig:refinement})
and 2) appropriate hanging-node constraints. For 3D, the implementation of the missing constraint definitions is in progress.
\item TODO
\end{itemize}

Furthermore, we continued to remove the instance of usage of the \texttt{GeometryInfo} class from
the library and to replace them by more general equivalent functions, which
rely on \texttt{ReferenceCell} in many cases. Once all instances of 
\texttt{GeometryInfo} are removed, we will deprecate the class.

@@

Until now, distributed meshes were either partitioned statically (\texttt{parallel::\allowbreak fullydistributed::\allowbreak Triangulation}; short \texttt{p::f::T}) or they used a
fixed policy to partition the cells among processes
(\texttt{parallel::\allowbreak distributed::\allowbreak Triangulation}; short \texttt{p::d::T}). For the latter, we
use the Morton-order as the backend of \texttt{p::d::T}---\texttt{p4est}---does. A fixed Morton-order
might provide very good performance in terms in communication and
setup costs, but might be non-optimal when interfacing with other
libraries that have fixed partitioning, e.g., a Cartesian
partitioning, themselves.

New utility functions from the \texttt{RepartitioningPolicyTools} namepace can be
used now to create a new \texttt{p::f::T} instance,
given a distributed triangulation (\texttt{p::f::T} or \texttt{p::d::T}) and a vector with the new owners of locally
owned cells. The workflow is shown in the following listing:
\begin{c++}
// original triangulation
parallel::distributed::Triangulation<dim> tria(comm);

@@

// create the new triangulation
parallel::fullydistributed::Triangulation<dim> tria_pft(comm);
tria_pft.create_triangulation(construction_data);
\end{c++}
Instead of using one of the predefined partitioning policies, users
can write their own ones by implementing the \texttt{RepartitioningPolicyTools::Base} interface. Similarly to the
active level, also the multigrid levels can be repartitioned arbitrarily.

\begin{figure}
    \centering
    \def\svgwidth{0.8\columnwidth}
    \input{svg/repartitioning.pdf_tex}
    \caption{Visualization of the repartitioning process: after new ranks 
    are assigned to cells, each process collects and sends the cells, incl. 
    ghost cells, to the new owner. There, incoming cells are processed, duplicates
    are removed, and the local part of the triangulation is built.}\label{fig:repartitioning}
\end{figure}

The setup process is visualized in Figure~\ref{fig:repartitioning}. At first, cells, incl. their
surrounding (ghost) cells and parent cells, are collected and 
sent to the new owner. On the
receiving side, the sets of all cells are combined and possible duplicates
are removed. This information is enough to construct a new triangulation.
For sending/receiving, we apply consensus-based algorithms~\cite{hoefler2010scalable}, which
we introduced into the library in release 9.2~\cite{dealII92}. Consensus-based 
algorithms are used also in \cite{ibanez2016pumi} for repartitioning.

In future releases, we plan to add support for repartitioning based on distributed graph
partitioning libraries, e.g., \texttt{ParMETIS} or \texttt{Zoltan}. Furthermore, we intend to extend 
\texttt{p::f::T} with adaptivity support.
With this support and and the new repartitioning features, \texttt{p::f::T}
could be self-consistent like \texttt{p::d::T}, which is an important
step towards the support of AMR also for distributed simplex and mixed meshes
(see also Subsection~\ref{sec:simplex}).

@@

In the matrix-free infrastructure, we have added a number of new features. The most
notable ones are:
\begin{itemize}
\item Improved support of Hessians: just like values and gradients, Hessians can be
now evaluated and integrated during matrix-free loops
both for cells and faces. This enables, e.g., to write
a matrix-free version of step-47, which solves the biharmonic equation with DG.
\item Cell-centric loops now also allow to access gradients and Hessians
of neighboring cells on faces. The major difficulty here was their reorientation
in the case of arbitrary orientated cells on unstructured meshes.
\item Users can now create their own cell batches, by providing \texttt{FEEvaluation::reinit()} a list of cell IDs. \texttt{FEEvaluation}
accesses the appropriate data and reshuffles mapping data accordingly on
the fly in order to enable vectorization over cells. The new feature is useful in many
contexts. Examples are application with sharp interfaces (e.g., two-phase flow
or shock capturing), where one needs to treat cells that are ``cut'' by
the interface in a special way. A challenge is that cell batches
might contain cut or not-cut cells, making vectorization potentially more complicated. One way to deal with such cell batches is to apply masks if
the code paths do not diverge too much
(a lot of functions of \texttt{FEEvaluation} allows this). Another way is to
categorize cells during \texttt{MatrixFree::reinit()} in such a way that mixed cell batches do not occur. However, calling  \texttt{MatrixFree::reinit()}
might be too expensive if a very dynamic system is given, which requires
recategorization in each time step. Doing this on the fly might be a cheep alternative, even if not computationally optimal during matrix-free loops.
\end{itemize}
Besides these new features, we improved the performance of 
hanging-node-constraint evaluation on the CPU. Instead of performing
quasi-dense matrix-vector multiplications~\cite{KronbichlerKormann2012}, we now use an
approach based on inplace interpolation and sumfactorization, similarly
as we already did in the GPU code~\cite{ljungkvist2017matrix}. In \cite{munch2022hn}, the algorithm is described 
and performance numbers are shown, indicating a reduction
of overhead of cells with hanging nodes by a factor of 10$\times$.

Furthermore, we have performed a major restructuring of internals
of the \texttt{FEEvaluation} classes. This will enable us to add the support of new element types, e.g., of 
Raviart--Thomas and Nedelec elements,  more
simply in the future.

@@

In release 9.3~\cite{dealII93}, we added support for global-coarsening multigrid in 
addition to the established local-smoothing infrastructure. Global 
coarsening smoothens on the whole computational domain on each
multigrid level, which is obtained by coarsening the finest mesh globally.
For this purpose, we use a sequence of triangulations, and we perform
the smoothing only on their active levels. To create the sequence of 
triangulations, one can use the functions \texttt{create\_geometric\_coarsening\_sequence()} from the
\texttt{MGTransferGlobalCoarseningTools} namespace. A new version takes an 
instance of \texttt{RepartitioningPolicyTools::Base} (see Subsection~\ref{sec:repartitioning}) as argument, which allows to specify the parallel 
distribution of each multigrid level (in contrast to the fixed first-child policy in the case of local smoothing). Furthermore, we added support for block vectors,
fixed a number of limitations, and performed performance optimizations of 
the transfer operator; particularly, the redundant copy from/to temporary vectors
has been eliminated. Now, hanging-node constraints are applied
efficiently in the same way as in the matrix-free loops (see Subsection~\ref{sec:mf}).

In \cite{munch2022gc}, the performance of the local-smoothing and global-coarsening
infrastructure of \texttt{deal.II} was compared for locally refined meshes. The results indicate that
the local definition of multigrid levels might introduce load imbalances
in the case of local smoothing so that global coarsening is favorable despite
potentially more expensive intergrid transfers. In order, to judge the benefits
of one approach against the other, \texttt{deal.II} provides new functions
\texttt{workload\_imbalance()} and \texttt{vertical\_communication\_efficiency()} 
in the \texttt{MGTools} namespace for the  estimation of the workload imbalance and the
vertical communication efficiency purely based on the given mesh. The publication \cite{munch2022gc} also points out that not
all types of smoothers are applicable for global coarsening due to the
presence of hanging nodes, which is a motivation to add new smoother types
to \dealii in the future.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{New and improved tutorials}
\label{subsec:steps}

@@

The 9.4 release includes
\href{https://dealii.org/developer/doxygen/deal.II/changes_between_9_3_0_and_9_4_0.html}
{around 40 incompatible changes}; see \cite{changes94}. The majority of these changes
should not be visible to typical user codes; some remove previously
deprecated classes and functions; and the majority changes internal
interfaces that are not usually used in external
applications. That said, the following are worth mentioning since they
may have been more widely used:
\begin{itemize}
  \item In continuation of our attempt to merge the classes \texttt{DoFHandler} and \texttt{hp::DoFHandler}, we have removed the
  template parameter \texttt{DoFHandlerType} from number of classes and
  functions and replaced it by \texttt{dim}/\texttt{spacedim}. Affected
  classes are, e.g., \texttt{SolutionTransfer} and \texttt{DataOut}. 
  \item \texttt{FE\_RaviartThomasNodal} now uses a different polynomial space to allow
for a simpler use for faces in non-standard orientation. The new polynomials
are anisotropic tensor products of Lagrange polynomials on the points of the
Gauss--Lobatto quadrature formula. This change leads to different entries, for example, in
the matrices and constraints, but no change in accuracy should be expected as the resulting polynomial
space spans the same polynomials.
\item We made different changes in the repartitioning of \texttt{parallel::distributed::Triangulation}. Most notably, we have
removed the default weight (10,000) in order to ensure consistency with the rest
of the library.
